% -*-latex-*-
% 
% For questions, comments, concerns or complaints:
% thesis@mit.edu
% 
%
% $Log: cover.tex,v $
% Revision 1.8  2008/05/13 15:02:15  jdreed
% Degree month is June, not May.  Added note about prevdegrees.
% Arthur Smith's title updated
%
% Revision 1.7  2001/02/08 18:53:16  boojum
% changed some \newpages to \cleardoublepages
%
% Revision 1.6  1999/10/21 14:49:31  boojum
% changed comment referring to documentstyle
%
% Revision 1.5  1999/10/21 14:39:04  boojum
% *** empty log message ***
%
% Revision 1.4  1997/04/18  17:54:10  othomas
% added page numbers on abstract and cover, and made 1 abstract
% page the default rather than 2.  (anne hunter tells me this
% is the new institute standard.)
%
% Revision 1.4  1997/04/18  17:54:10  othomas
% added page numbers on abstract and cover, and made 1 abstract
% page the default rather than 2.  (anne hunter tells me this
% is the new institute standard.)
%
% Revision 1.3  93/05/17  17:06:29  starflt
% Added acknowledgements section (suggested by tompalka)
% 
% Revision 1.2  92/04/22  13:13:13  epeisach
% Fixes for 1991 course 6 requirements
% Phrase "and to grant others the right to do so" has been added to 
% permission clause
% Second copy of abstract is not counted as separate pages so numbering works
% out
% 
% Revision 1.1  92/04/22  13:08:20  epeisach

% NOTE:
% These templates make an effort to conform to the MIT Thesis specifications,
% however the specifications can change.  We recommend that you verify the
% layout of your title page with your thesis advisor and/or the MIT 
% Libraries before printing your final copy.
\title{Driverless Dreams: Technological Narratives and the Shape of
  the Automated Car}
%% Replacing or Augmenting the Human
%% Chauffeurs or Cyborgs: Ideologies and the Shape of the Self-Driving Car
%TODO change the title of my other piece if I go with this thesis
%title
%Self-Driving Stories: Automated Vehicles and the Human Being
%The Shape of the Self-Driving Car: Ideologies, Strategies, and the
%Human Being
%Vehicles of Ideology: Narratives, Impacts, and the Shape of the
%  Self-Driving Car
%Self-Driving Dreams: Narratives, Ideologies, and the Shape of the
%Automated Car

% Driverless Dreams: Narratives, Ideologies, and the Shape of the Automated Car
% Asleep at the Wheel: Dangerous Fantasies of a Driverless Future



\author{Erik Lee Stayton}
% If you wish to list your previous degrees on the cover page, use the 
% previous degrees command:
       \prevdegrees{Sc.B. Physics A.B. English, Brown University (2007)}
% You can use the \\ command to list multiple previous degrees
%       \prevdegrees{B.S., University of California (1978) \\
%                    S.M., Massachusetts Institute of Technology (1981)}
\department{Program in Comparative Media Studies/Writing}

% If the thesis is for two degrees simultaneously, list them both
% separated by \and like this:
% \degree{Doctor of Philosophy \and Master of Science}
\degree{Master of Science in Comparative Media Studies}

% As of the 2007-08 academic year, valid degree months are September, 
% February, or June.  The default is June.
\degreemonth{June}
\degreeyear{2015}
\thesisdate{May 8, 2015}

%% By default, the thesis will be copyrighted to MIT.  If you need to copyright
%% the thesis to yourself, just specify the `vi' documentclass option.  If for
%% some reason you want to exactly specify the copyright notice text, you can
%% use the \copyrightnoticetext command.  
%%\copyrightnoticetext{\copyright Erik Stayton, 2015. All rights reserved.}

% If there is more than one supervisor, use the \supervisor command
% once for each.
\supervisor{T. L. Taylor}{Associate Professor of Comparative Media Studies}
\supervisor{David A. Mindell}{Professor of the History of Engineering and
Manufacturing (STS) \\ Professor of Aeronautics and Astronautics}

% This is the department committee chairman, not the thesis committee
% chairman.  You should replace this with your Department's Committee
% Chairman.
\chairman{T. L. Taylor}{Department Committee Chair, Director of Graduate Studies}

% Make the titlepage based on the above information.  If you need
% something special and can't use the standard form, you can specify
% the exact text of the titlepage yourself.  Put it in a titlepage
% environment and leave blank lines where you want vertical space.
% The spaces will be adjusted to fill the entire page.  The dotted
% lines for the signatures are made with the \signature command.
\maketitle

% The abstractpage environment sets up everything on the page except
% the text itself.  The title and other header material are put at the
% top of the page, and the supervisors are listed at the bottom.  A
% new page is begun both before and after.  Of course, an abstract may
% be more than one page itself.  If you need more control over the
% format of the page, you can use the abstract environment, which puts
% the word "Abstract" at the beginning and single spaces its text.

%% You can either \input (*not* \include) your abstract file, or you can put
%% the text of the abstract directly between the \begin{abstractpage} and
%% \end{abstractpage} commands.

% First copy: start a new page, and save the page number.
\cleardoublepage
% Uncomment the next line if you do NOT want a page number on your
% abstract and acknowledgments pages.
% \pagestyle{empty}
\setcounter{savepage}{\thepage}
\begin{abstractpage}
\input{abstract}
\end{abstractpage}

% Additional copy: start a new page, and reset the page number.  This way,
% the second copy of the abstract is not counted as separate pages.
% Uncomment the next 6 lines if you need two copies of the abstract
% page.
% \setcounter{page}{\thesavepage}
% \begin{abstractpage}
% \input{abstract}
% \end{abstractpage}

\cleardoublepage

\section*{Acknowledgments}

Any substantial work of writing has a long gestation period, and this
has been no exception. A great number of people come to mind who have
contributed to the existence of this thesis over the last 20 months,
but I fear I must 
inevitably leave some out. To them I can say only \emph{mea maxima culpa}. 

No master's thesis is complete without a
solid advising team, and my most heartfelt thanks go to T. L. Taylor
for her careful reading and editing, and her ceaseless dedication (including periodic
game nights) to ensuring we all make it through the master's process with most of our
sanity intact. David Mindell has also been a continual resource
throughout the writing of this thesis---and indeed is cited frequently
within it---and I look forward to continuing to work with him on
future projects. Nick Montfort (who has exhorted me repeatedly to, as
much as possible, interrogate the technical objects themselves),
William Uricchio (who has challenged my terminological precision),
Heather Hendershot (who has kept my eye on science fiction) have all
been indispensable in their own ways, and have given freely of both
thesis and career advice. Discussions with Kate Crawford, Nick Seaver,
Tarleton Gillespie, Merritt Roe Smith, Stefan Helmreich, and G\"{o}de Both have also
shaped this work and made it what it is; as have innumerable
discussions at conferences. Conversations with a number of informants
(including Ryan Chin, Jim Womack, and John Leonard)
furnished important material, and though I cannot list all of them by
name, I wish to thank them all for giving of their time.

All of my classmates, in the classes of 2014, 2015 and 2016 have been great
friends, supporters, and sounding boards for ideas. But I fear I must
specifically thank Liam Andrew, Chelsea Barabas, Suruchi Dumpawar,
Jesse Sell, Ainsley Sutherland, and Wang Yu for discussing
self-driving vehicles and technological ideology with me 
more times than some of them probably cared to---and additionally Desi
Gonzalez, Kyrie Caldwell, and Lacey Lord for being available for
games and gripe sessions throughout the process. On that note, my old
friend Kristin Bergman, and my new friend Andrea Morales Coto, have
been invaluable companions on this adventure, providing much needed
levity and perspective in trying times. I also thank my forthcoming
classmates at MIT HASTS for their input and encouragement,
particularly Alison Laurence and Clare Kim.

My parents, Barbara Mindell and Lee Stayton, have often been the first
to hear my ideas, and have provided irreplaceable editing and feedback
on this and other projects during my two years at MIT. I am so pleased
to have been able to share this work with you, and both it, and I,
have profited immensely from your input and your presence. Abigail
Strait, whose food, hospitality, and seemingly endless supply of tea I have
partaken of innumerable (at least more than twenty!) times during the
writing of this thesis, you will always have my gratitude.

But most of all, Evan Donahue, whose keen commentary, acerbic wit, and
drive to re-envision the future of humanity has spurred me on: this
project exists in no small part due to his influence, and he most of
all deserves thanks. As we have 
often said, this was supposed to be the Fourth Great and Bountiful
Human Empire. Let us strive make it so.

 

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% -*-latex-*-
