\chapter{Introduction}
\label{chap:0}

%Intro / Visions / Public Faces (14 p)

%%A FEW IMAGES for this chapter

It is early August, 1925, and a strange spectacle greets viewers on
Broadway in New York City. An empty car lurches uncertainly down the
street,\footnote{This description is adapted from \emph{Time
 Magazine} \cite{ScienceRadio}.} followed closely behind by a
chase car, filled with radio equipment and several men, one driving,
and one furiously operating a set of controls. The lead car features a
diamond-shaped antenna sprouting from where the back seat should be,
which is instead filled with a motley collection of dials and tubes,
wires and batteries. A man stands on the running board of the
jittering contraption, but keeps his
hands well away from the controls, as if to tempt fate, as the car,
unsteady in its movements, turns onto Fifth Avenue and comes,
ungracefully, to a halt.

\section{Concepts, Models, Dreams}

While present work on automated vehicles by Google, Mercedes, Tesla,
and others might lead one to think that driverless vehicles are a
new idea, specifically enabled by recent advances in artificial
intelligence, they have not merely been dreamed about, but actually
built in prototype form, as far back as 1925, when Houdina Radio Control
built the ``Linrrican Wonder'' from a 1926 Chandler, and demonstrated
it live on the streets of New York City. But it would be wrong to
characterize a 90-year engineering journey as the continual
progression toward a ``modern,'' developed device. There is nothing
fundamentally natural about the way automated---or, often interchangeably
``autonomous,'' ``driverless,'' or ``self-driving''\footnote{I indeed use these terms
  interchangeably at times even within this text, so I should mention
the specifics of their use here. Autonomous, driverless, and self-driving
vehicles are generally one and the same, though the terms
``driverless'' and
``self-driving'' are used particularly in the press when talking about
these systems. These all presuppose largely self-contained operation
on the part of the automated system. An ``automated'' vehicle however
does not necessarily presuppose self-contained operation, and cars
have become increasingly automated over the past 50 years (through
ABS, traction control, cruise control, automatic transmissions, etc.) without, so
far at least, becoming autonomous. This point, simple though it
sounds, is at the center of this thesis. I use \emph{automated} to refer to any amount
or kind of automation, and the other terms interchangeably when
implying more complete automation, or conventional ideas thereof.}---vehicles are being
envisioned, designed, and talked about. Instead, automated vehicle
development has sustained several successive, overlapping paradigms,
in part, though not completely, driven by the technical capabilities
of and excitement about contemporary technologies. However, equally
important in this story are the ideologies of control and
human-machine interaction that shape engineering practice and vehicle 
development, and alter how developments are represented and discussed
in popular media. What we believe self-driving cars to be---or what we
believe they soon will be---is a product of the popular, teleological understanding
of technology development, which draws from existing technocratic
narratives. Alternative paradigms exist,
but deserve much broader recognition outside of certain engineering circles. In
order to understand the way that automated vehicles have been figured
popularly, the implications of that vision, and the alternatives to
it, we should start by examining the varied types and dimensions
of historical automotive automation research.

%% \footnote{These narratives are deeply, though 
%%   not exclusively, American, and are intertwined with stories of
%%   American technological progress over the last century, from the
%%   Manhattan Project to the Apollo Program to ARPAnet to IBM's
%%   Watson.}

%% It is
%% vital that we recognize t

Self-driving cars are born from the
same history as interchangeable parts, assembly lines, machine tools,
and scientific management; as electrification, computerization, and
networked interconnection. When asked how to make something better,
safer, faster, easier, more productive, we involve machines or at
least what H. M. Collins describes as ``machine-like actions''
\cite[p. 42]{Collins}. This is not a new predilection, but is
something that has been with us since the Industrial Revolution.
Furthermore, technological visions and concepts inscribe excitement about the
technologies of the present moment, and this tendency has deeply
affected the history of automated vehicle research. From the railroad to the
Internet (passing through the telegraph, the airship, electrification,
television, the automobile, the rocket), each new mode of
communication (or transportation\footnote{Rudolf Arnheim referred to
  TV, a ``relative of motorcar and airplane,'' as a ``means of
  cultural transportation'' \cite[p. 194]{arnheim}.}) ushers in new visions of our
collective future. We are famously bad at these predictions \cite{vanriper}, but they
nevertheless drive the popular imaginary, which has included automated
vehicles since the 1920s. Both visions of and research on self-driving
technology have taken on aspects of the central information
technologies of their day, and re-inscribed them as control
technologies for physical devices. The history of our driverless
dreams shows that the form of automated vehicles has been neither
constant nor a foregone conclusion.

\subsection{Early Work}

Houdina's early experiments in radio-controlled vehicles (which began
this chapter) put us into a time period in which radio was an
exciting, emerging technology that was envisioned as a way to take
the driver out of the vehicle, even if a human's direct commands were
still necessary for vehicle operations. Aachen Motors's ``Phantom
Auto,'' operating on the same principles, toured Milwaukee a year
later, \cite{phantomAuto} and returned as an attraction in
Fredericksburg in 1932 \cite{phantomAuto2}.\footnote{Radio control had
  at this point been public knowledge for at least two 
decades, since Nikola Tesla's 1898 demonstration of a radio-controlled
boat, but though it was not fundamentally new, the use of the
technology on public roads nevertheless shocked onlookers.} Contemporary news articles
describe the machine as if it drove itself, 
as if the physical position of the person was what mattered, not the
fact that a human was always using the radio transmitter:  ``Driverless,
it will start its own motor, throw in its clutch, twist its steering
wheel, toot its horn, and it may even `sass' the policeman at the
corner'' \cite{phantomAuto}. Described as ``one of the most amazing products of
modern science,'' the 
car proceeds ``as though there were an invisible driver at the wheel''
in an effect both ``uncanny and mystifying'' \cite{phantomAuto2}. Such work,
though it seems relatively primitive in retrospect (hardly more than a
glorified RC-car, though it removed the driver from a full-size vehicle), may well have
helped capture the popular imagination. This compelling idea would reemerge
a few years later.

%% , one which included trains and aircraft, a farm, and an
%% amusement park
%% For their concept cars, GM separated the control
%% functionality for steering from that for the maintenance of speed and
%% distance between cars.
%% \footnote{As a
%%   side note, this has been described not only as techno-utopian but as
%% an outgrowth of commodity scientism, a triumph of a sort of
%% marketing-driven approach to technology that is more rhetoric than
%% substance \cite{smithSelling}It seems likely that commodity
%% scientism, or a similar phenomenon, is deeply involved in the ways
%% modern technological narratives are generated and consumed.}

%% that would restore the position of the cars using
%% gravity, a low-tech solution to the more complicated problem of
%% steering. However,

The 1939 New York World's Fair opened to great acclaim. Covering over
a thousand acres and attracting a total of over 44 million visitors,
the fair was a massive public spectacle demonstrating the technical
prowess of American industry and providing a grand vision for the
future of the nation. The structure of the exhibition---exemplified in
its grand (and oddly-named) buildings, the Perisphere and
Trylon---manifested the modernist techno-utopian ideology that drove the
dramatic technological displays \cite[p. 371]{nyeElectrifying}. The
great success of the fair was 
GM's Futurama exhibit\footnote{The line for the exhibit routinely stretched to
two miles in length, and 28,000 people visited the exhibit each day.}
which presented an idea of what the world would look
like in 1960 \cite[p. 3-4]{wetmore}. Automobiles were the focal
point: the bird's-eye view presented a 
future city with multi-lane elevated expressways filled with largely
automated vehicles. While the cars' lateral position was maintained via
curved barriers, the Futurama exhibit described that distance
between cars would be maintained via a ``sophisticated system of radio
control'' \cite[p. 5]{wetmore}.\footnote{Prior to electronic circuits, this would have been
accomplished with vacuum tubes, a technology well-known at the time,
and certainly subject to significant hurdles in terms of mechanical
reliability. It was not until digital electronics were invented that
hardware became the relatively stable component of driverless systems,
with software reliability becoming the key issue \cite[p.
  15]{wetmore}.} Despite the 
fragility of glass vacuum tubes traveling at high speed, and despite the fact
that the technology could work well only for sufficiently
well-designed roads (high-speed highways, built specifically for these
vehicles, with attendant infrastructural costs), such a system was
reasonably coherent as a vision of what 1960 might be, from the
perspective of attendees. This aspirational future was clean,
modernist, and efficient; the rationalization of the
factory\footnote{Public interest in this topic was demonstrated through
  popular factory tours of major Detroit automotive 
  plants, and 
working GM and Ford assembly lines at the earlier Century of Progress
Exhibition in Chicago (1933-1934) \cite[p. 63-64]{nyeAmericas}.} had
spread to the city and 
countryside, where working hours were reduced through electrification,
and even driving could become leisure, the mere inhabitation of a
device operating via radio control.

Radio was not the only revolutionary medium to be included in
 portrayals of the self-driving car of the future. The cover of the
 April 1936 edition of the magazine \emph{Modern
   Mechanix} \cite{modernmechanix} betrays a 
 different vision of the autonomous vehicle: the so-called ``electric
 eye automobile'' which would steer itself via a control loop, using an
 array of photocells on the car to track a light
 beam projected from the car and reflected by mirrors in the road
 surface.\footnote{Intriguingly, this simple application has become
   one of the classic robotics teaching experiments. I remember
   building similar robots in high school robotics club.} As the article, titled ``Light Beams Steer Super Racing
 Cars,'' describes:
\begin{quote}
With speeds, such as recently attained by the famous Sir Malcom
Campbell, already approaching the point where human reflexes are too
slow to insure safe control of the car, science has turned to the
photo electric cell for a possible solution. A proposed driverless car
involves the use of multiple electric eyes as the heart of its
steering mechanism. A powerful beam of light directed at a large lens
on the front of the car is concentrated on steel mirrors set at an
angle in the trackbed. The reflections are ``caught'' by the electric
eyes which convey the electrical impulses to a mechanical-electrical
brain which keeps the speeding car on its course \cite[p. 71]{lightbeams}.
\end{quote}

The cover pictures an intrepid ``driver'' using a camera to take a
``movie recording'' of the car's performance, while he leaves the
``driving responsibilities to the mechanical and electrical brain''
\cite[p. 71]{lightbeams}.
Responsible for initially putting the car into motion, the driver can
then step back from the task of driving, allowing the electrical
controls of mechanical linkages within the car---meticulously, if only
artistically, diagrammed
within the magazine article---to convey him or her safely and speedily down
the track. The photocells within the vehicle's sensing mechanism would
operate relays controlling the steering linkage, closing the loop
between sensing and acting in the manner of a ``teleological''
\cite{behaviorPurpose}, 
self-governing mechanism with corrective feedback:  the simple
idea of feedback control opens a vast array of new potentialities for
the future. But how 
the car's governor could be modified to allow it to race
beside other cars is not mentioned, which is telling since the
complexities of the environment are what make automated driving so difficult.

\subsection{The Midcentury}

%% ---in this case
%% the vision of flying cars shown in \emph{The Jetsons} in 1962, a
%% dream related to, but still distinct from, the driverless dream--

%% Matt Novak\footnote{Novak writes the
%%  blog Paleofuture, once on the \emph{Smithsonian} website, now on \emph{Gizmodo},
%%  about past visions of the future.} suggests that

By 1960, driverless vehicles had not been brought to market, but
interest in the technology continued. Ford's vice president of
engineering and research, Andrew Kucher, was referenced in the \emph{Chicago
Daily Tribune}, April 25, 1959, in a speech he gave at Northwestern
University, taking seriously the idea of autonomous cars. The article,
titled ``In 50 Years: Cars Flying Like Missiles!'' asks readers: ``Can
you imagine flying automobiles directed by automatic guidance
systems?'' \cite{tribuneViaPaleo}. ``Arthur Radebaugh's syndicated
Sunday comic `Closer Than We Think' was also a likely inspiration''
for contemporary images of advanced car technologies (\emph{The Jetsons}), as in
1958 it depicted hovering 
cars floating on an air cushion in an ``already proved'' concept
publicized by the same Andrew Kucher \cite{novakRecapping}. Newspapers
from April of that 
year also describe a 3-foot-long model of Kucher's ``Glideair'' that was
demonstrated to reporters in Detroit \cite{novakFlying}. The car may
not be driverless, 
but the people depicted riding in the ``flying carpet car'' in
Radebaugh's illustration do not seem to be paying much attention to
where it is going, and the road surface is bordered with ridges that
visually suggest the idea that the car will keep itself within the
demarcated space.

%% and the thoroughly contemporary
%% dashboard and instrument cluster,

One particularly salient public image from the 1950s comes from the
\emph{Saturday Evening Post}, which ran an advertisement for America's
Electric Light and Power Companies that depicted self-driving cars
coasting through an idyllic landscape, guided by electricity
\cite{weber}. The roadways 
are clean and clear, stretching off confidently into the distance. The
landscape, tamed and controlled, yet retains its natural beauty:
nowhere to be seen are smokestacks and the power plants generating the
guiding force of these vehicles. The rural, even
pastoral, landscape is largely unbesmirched, only cut across by the
rationalization of a futuristic interstate highway system.\footnote{Recall that
the Federal Aid Highway Act of 1956 began the creation of the
interstate system as we have it today.} Within the finned, bubble
canopied auto sits a nuclear family---father,
mother, daughter, and son---enjoying family time.
Perfectly coiffed, snappily dressed, and totally at ease, the mother
and daughter play dominoes, while the son admires a model of a futuristic
delta-wing jet aircraft and the father looks on. Their faces glow with
expressions of domestic contentment. The father, in the ``driver's
seat,'' is turned away from the wheel, facing backward. The caption reads: 
\begin{quote}
One day your car may speed along an electric super-highway, its speed
and steering automatically controlled by electronic devices embedded
in the road. Travel will be more enjoyable. Highways will be made
safe---by electricity! No traffic jams ... no collisions ... no
driver fatigue \cite{weber}.
\end{quote}
The limitless possibilities of electricity, measured and controlled by
America's Electric Light and Power Companies, reproduced a conservative
present into an equally conservative future. 

%% If electricity was still the force that would be harnessed to make the
%% roadways safe, what were the ``mechanical equivalents of sense
%% organs'' \cite{wienerMachineThreat} for the driverless car? 
%% But unlike the 
%% light-guided vehicle from the 1930s, this car had a new guidance
%% mechanism, which is entirely 
Wartime and postwar developments in electronics, the same ones that
inspired new science fiction visions of the future, drove
research in automated vehicle systems sensing and control. In 1953, General
Motors and RCA completed a scale model of an automated highway system
that would allow the two companies to experiment
with developing steering and distance control systems \cite[p. 6]{wetmore}.
General Motors's 1950s Firebird concept
vehicles marketed themselves on this research, despite the fact that
Firebird II ``had no automated capabilities whatsoever'' \cite[p. 7]{wetmore}. GM's
promotional films suggested the vehicle could be controlled
electronically from traffic control towers---like those in use in
aviation---placed along major highways: the car was ``under the direction
of an `electronic brain' on a dream highway of the future'' \cite[p.
  7]{wetmore}. This 
would have been the realization of the New York World's Fair exhibit
from two decades earlier. Building to make
this optimistic vision of GM's engineering ability a reality,
engineers led by Joseph Bidwell and Lawrence Hafstad within GM
Research installed ``pick-up coils'' on a 1958 Chevrolet: through a
feedback system like that of the ``electric eye automobile,'' these
coils would detect a wire embedded in the road and adjust the
vehicle's steering \cite[p. 7]{wetmore}. Consistent with the image presented in
the \emph{Saturday Evening Post}, and a significant advance from
Futurama's mechanical ``half-pipes,'' alternating current would
be the control mechanism of the future. A GM press release announced:
\begin{quote}
An automatically guided automobile cruised along a one-mile check road
at General Motors Technical Center today, steered by an electric cable
beneath the concrete surface. It was the first demonstration of its
kind with a full-size passenger car, indicating the possibility of a
built-in guidance system for tomorrow's highways.... The car rolled
along the two-lane check road and negotiated the banked turn-around
loops at either end without the driver's hands on the steering wheel
\cite[p. 7]{wetmore}.
\end{quote}
 Also in the 1950s, RCA's
Vladimir Zworykin, a lead inventor of television technology, was
working on an intelligent road system of his own. His concept,
inspired by ``railroad block signals,'' used circuits embedded in the
road to magnetically sense vehicle speed and location, placing sensing
and coordination capabilities outside of the vehicle:  Zworykin's
centralized planning model would send 
instructions to individual cars, and a 1/40th scale demonstration
system was built for the 1960 Highway Research Board meeting in
Washington D.C \cite[p. 9]{wetmore}.

%% Instead of the mechanical ``half-pipes'' of the 1939 Futurama roadway, advances in
%% technology would allow a future of automated steering powered
%% by alternating current. Electromagnetism supplemented electricity as the
%% guidance mechanism of choice.

%% (who gave numerous
%% public lectures through the 1950s) 

The 1964 New York World's Fair hosted a second Futurama exhibit, in
which GM presented an automated highway system much like Zworykin's
system next to multitudinous nuclear-powered concepts. The exhibit was
thus described: 
\begin{quote}A revolutionary `Autoline' expands the capacity of a three-lane
expressway: Electronically, a control-tower operator steers, brakes
and sets the speed of each car in an automatic lane; groups of cars
move at equal intervals as a group \cite[p. 9]{wetmore}.\footnote{This
idea may seem quaint, but actually represents an important operations
paradigm that is the focus of the last two chapters of this thesis.
Public proposals to this effect are rare, but the concept remains
important.}\end{quote}
 Advances in
electronics, and the expansion of early control theory into
cybernetics through the work of Norbert Wiener and his contemporaries seemed
poised to make revolutionary applications of electrical control
possible. Concern about human reaction times persisted. In a lecture given in 1960, Wiener
make the case for automated control systems: ``by the time we are able
to react to our senses and stop the car which we are driving, it may
already have run head on into a wall,'' he wrote, and the answer to
this problem was the lightning speed of cybernetic control \cite{wienerOn}. But real
developments in automated driving did not materialize, and research
languished, out of the public eye, throughout the late 1960s and 1970s
\cite[p. 10]{wetmore}. Instead of autonomous vehicles---in a trend
that arguably began
with \emph{Safety Last} \cite{oconnellmyers} and \emph{Unsafe at Any Speed} \cite{nader}---the public pushed
for an
increasing array of safety systems designed to make vehicles more forgiving
of human errors---starting with rudimentary
mechanical restraints such as seat belts, and continuing through
airbags, ABS, and other more complex automation systems.

\subsection{Research at the Turn of the Century}

%% FOOTNOTE the acronyms
%%  Typical of vehicles of the time,
%% VITA\footnote{VIsion Technology Application. The original VITA was a
%%   D811 Mercedes van, chosen in order to store both computing
%%   equipment, power supplies, and environmental control for those
%%   electronics \cite[p. 37]{ulmerVITA}.} used
%% an analog video-camera signal processed through a framegrabber
%% which would feed the visual information into the transputer-based
%% parallel-processing system \cite[p. 38]{ulmerVITA}.

Not all vehicle research focused on cars themselves, and active
efforts to design and develop intelligent vehicle-highway systems (IVHS) occurred
in the United States and elsewhere during the 1980s and 1990s \cite[p.
  11-12]{wetmore}. These
efforts focused primarily on ``nearer term'' technologies, such as
electronic tolls and driver information systems,\footnote{Such as 511
  systems \cite{costello511}.} located
outside the vehicle. In contrast to earlier radio or track-based
electrical control, which similarly involved infrastructural changes, the 
``modern'' era of self-driving vehicle research begins with
computer-vision based approaches. While computer vision techniques had
been experimented with for vehicle guidance since 1969, it was not
until the 1980s that microelectronics became powerful enough to
process images in near real time and compact enough to be placed on
the vehicle itself \cite[p. v]{dickmannsDynamic} \cite{vamors-p}. In the 1980s, Ernst
Dickmanns's lab at the 
University Bundswerhr, in Munich, Germany was active in some of the
earliest self-driving car research and development. Their early
vehicle, a specially equipped 5-ton Mercedes-Benz van---computerized
controls could be used to perform all necessary driver inputs---was
fitted with cameras, other sensors, and an image-processing system to
close the control loop and drive the vehicle based on visual
information \cite{vamors-p}. This early experiment,
VaMoRs,\footnote{Versuchsfahrzeug f\"{u}r autonome Mobilit\"{a}t und
  Rechnersehen, or ``test vehicle for autonomous mobility and computer
  vision''} paralleled in the United  
States with a Carnegie Mellon-developed vehicle based on a Chevrolet
panel van \cite{bogostHist}, was successfully tested on roads without
traffic at up to 
60 miles per hour, and was soon followed by new projects produced as
part of the EUREKA PROMETHEUS\footnote{PROgraMme for a European
  Traffic of Highest Efficiency and Unprecedented Safety} project
\cite{eureka}. 

The largest driverless vehicle 
research project of the time, funded by EU member states to the tune
of almost 750 million Euros, PROMETHEUS ran from 1987 to 1995 and
involved the VaMP\footnote{VaMoRs Passenger Car} vehicle from
Dickmanns's research lab along with its 
Daimler-Benz twin, VITA-II.\footnote{VIsion Technology Application. The original VITA was a
  D811 Mercedes van, chosen in order to store both computing
  equipment, power supplies, and environmental control for those
  electronics \cite[p. 37]{ulmerVITA}.} Like its forebear VITA, the VITA-II
system, scaled down to fit in 
a passenger car (an S-class SEL 500), used digitized analog
video signals to detect lanes and other vehicles \cite[p.
  2]{ulmerVITA-II}. Additional sensors 
detected brake pressure, temperature, steering angle, acceleration in
lateral and longitudinal directions, and yaw \cite[p.
  2]{ulmerVITA-II}. The camera systems
utilized two-camera setups with stabilized two-axis rotation to allow
them to follow objects of interest in the scene \cite{cameraPlatform}. And the
technology largely worked, with the PROMETHEUS project coming to a
successful conclusion in 1994 and 1995 with 1000 kilometers of largely
autonomous operation in normal traffic conditions on Paris motorways,
as well as a finale drive from Munich to Copenhagen \cite{makingBertha}.
Meanwhile, vision-based research also occurred actively in the United States;
Congress's passing of the Intermodal 
Surface Transportation Efficiency Act in 1991 allowed for \$650 million
in funding for driverless vehicle research over the following six
years
\cite{novakAlmost}.\footnote{Novak suggests that the biggest problem with
the legislation was that it failed to define what a ``fully automated
highway system'' would be, such that it could be achieved in the time
allotted \cite{novakAlmost}. This is an important consideration for
future work.} The act spurred a consortium of nine organizations to
work on meeting the 1997 deadline for demonstrations. In contrast to
IVHS, which was seen as fundamentally limited by the human driver,
this automated driving effort promised computationally-driven
improvements in safety and environmental impact, as Rodney Slater of
the Federal Highway Administration described in a 1993 announcement
\cite[p. 30]{wetmore}.
Notably, a Congressional report described the goal as ``hands-off,
feet-off'' driving \cite{novakAlmost}, language which returns in the
National Highway Traffic Safety Administration's recent
\emph{Preliminary Statement of Policy Concerning Automated Vehicles}
and proves similarly unhelpful.

%% , as will be described in chapter
%% \ref{chap:3}

%TODO(DONE) expand
%% Brief section about this project
%% note the prevalence of ``'hands-off, feet-off''' language, which
%% returns in NHTSA report in \ref{chap:1}
%% and specifically
%% Novak claims ``perhaps the biggest problem was that the legislation
%% never clearly defined what was meant by 'fully automated highway
%% system.''', an important consideration for future work

Further autonomous vehicle research has progressed from
these vision-based beginnings. The
DARPA Grand Challenge in 2004 and 2005 (evolving from the earlier
  DARPA Strategic Computing project), and the DARPA Urban Challenge 
in 2007, spurred significant interest, and brought together industry
groups and universities to 
try to extend the capabilities of fully autonomous vehicles.\footnote{The
military should not be discounted as a contributor to and funder of
  this research, and military visions of automated vehicles---as a way
  to remove human soldiers and transport drivers from harm's way---are
  sources of inspiration that deserve fuller
  exploration elsewhere.} While the
Grand Challenge technologies bear little resemblance to current
navigation approaches designed to deal with the complexities of
real-world road environments,\footnote{These vehicles basically
  followed a predetermined GPS path, with gross navigation skills to
  avoid obstacles at smaller scales. However, the desert environment,
  though hostile in many ways, is not nearly so complex as an urban or
suburban road. The navigational approaches applied there are
fundamentally different than those necessary for an every-day
self-driving vehicle. John Leonard,
  discussion with the author, December 3, 2014.} the Urban Challenge
vehicles, both via their sensor setups and algorithmic approaches to
navigation, can be interpreted as another step along the path to the
modern self-driving car, as instantiated within Google's design approach. 

\subsection{Automated Cars in 2015}

%% Even Apple is rumored, as of the time of writing,
%% to be working on an autonomous vehicle, though its approach has not
%% yet been made public \cite{taylorApple}.\footnote{Notwithstanding an
%%   unnamed industry source, there is some
%%   speculation that observed Apple-registered vehicles are a publicity stunt or part of
%%   a mapping exercise instead \cite{bonningtonCurious}.}

Approaches today have a variety of forms. While Google is making
headlines with its self-driving vehicle research, and is defining the
look of the autonomous car with its roof mounted LIDAR\footnote{LIght
  Detection and Ranging} scanner, other
companies are developing vehicles with slightly different sensors and
techniques. Google supplements cameras and LIDAR with detailed pre-mapping and highly accurate
differential GPS \cite{guizzoHow}. Mercedes, so far, focuses more on
cameras and radar, 
without using expensive LIDAR systems---and some of these features are
already seen in the new S-class luxury sedan, though these vehicles
have not yet reached the level of capability of Mercedes's
test vehicle \cite{makingBertha}. But these sensing choices come with their own
trade-offs in terms of operations, safety, and how safety can be
determined and proven. Tesla Motors
is also working on an iterated 
approach to vehicle autonomy, focusing publicly on software updates to
existing cars that add new automated features. Volvo is both involved
in a public test of vehicle autonomy in Sweden \cite{volvoPR}, and is moving forward
with automated systems in its production cars, building from years of
research in Advanced Driver Assistance Systems (ADAS) and Advanced
Vehicle Control Systems (AVCS). It is important to recognize that
a spectrum of approaches are being developed today, from iterated
improvements on current designs to radical approaches that attempt to
achieve full self-driving at once. Automated\footnote{I use ``automated''
  precisely here. It is one of the main failures of the
  ``self-driving'' term that it appears to make a clear distinction: 
  either something can drive itself or it cannot. But there are
  levels even here: Under what circumstances can it be self-driving? Is
  complete self-driving capability necessary before the term applies?
  These questions will return again and again through this thesis. }---or even
``self-driving''---vehicles are not just one, uniform type of
device. And, critically, there is no clear
line to be drawn between their extremes.


%% >>>Story hook: currently, this calls out to be a brief historical
%% sweep through self-driving car dreams from the 1920s/30s to the
%% present; but it could also begin with the reveal of the Google car; or
%% the Youtube video of the guy in the Mercedes tricking the lane assist
%% function to drive for him on the highway (2 p)
%% 1) establish what the technology is: in its varied types and
%% dimensions: this is not ONE THING or ONE IDEA (4 p)
%% ---INCLUDING historical models: see Erkki Huhtamo, media archaeology
%% (2 p)
%% >>>>pull from my essay!
%%THIS STUFF DUPLICATES CH1; sort it out!


%% DO include the CES depictions; and contrast with the 1950s ad, at
%% least in a footnote!
%% ---outside of discussions of
%% Google's desire to eliminate steering wheels facing off against
%% regulators, or questions of whether the automakers' gradual approach
%% or Google's ``moonshot'' approach will triumph
%% . In fact, the interior is
%% covered in touchable displays,
One would be excused, however, for thinking differently.
Portrayals of self-driving technology are often lacking in nuance. While
there is a fair amount of popular argument about whether these
vehicles will be beneficial (or how beneficial they will be) along
various axes of interest, there is relatively little discussion of
\emph{what} these vehicles will actually be: the implicit
assumption is that everyone is headed for the same technological
goals, the same ultimate type of device, approached gradually or all-at-once.

%%  (like the seating arrangement from the 1950s
%% \emph{Saturday Evening Post} ad)

At the Consumer Electronics Show in Las Vegas, in January
2015, visions of the driverless car stood front and center. Prototypes
and concepts at CES worked glitz and glamour into the dream of a
driverless future. Mercedes-Benz's F015 autonomous car
concept forgoes the traditional seating arrangement in favor of
rotating chairs that turn the center of the car into a lounge or
meeting space \cite{riofrioMB}. The car's side windows are relatively minimal, its
doors taken up primarily by touch screens, set up for an immersive digital
experience. But beneath this shiny media exterior sits a relatively
unglamorous purpose:  its publicity photos show serious-looking, young, white
businesspeople in uniform grey work clothing \cite{simpsonDesign}. The environment is
high tech and sharply clinical. Far from an exuberant depiction of the
promise of media technology in the automobile, this future is so
serious as to be dull: a homogeneous work space bleeding out into
other parts of life. Compared to America's Electric Light and Power
Companies' ad from the 1950s, where the car was a space for idyllic,
family life, what we think we will use these prospective vehicles for
has changed to the point where productivity is the stated or unstated
goal. Our assumptions about use have changed (with seemingly little
investigation into why), but they continue to be the background of both
optimistic and pessimistic viewpoints.

\section{Short-term Promises}

%reduce and deduplicate

One popular vision of this technology is fundamentally optimistic: it
puts forward the
idea that self-driving vehicles will be commercially available in as little
as five years, and that in just two decades the majority of cars on the road will be
fully autonomous. Even respected business-information and consulting
bodies have bought into this dream.\footnote{For example,
  IHS Automotive
  predicts 54 million such vehicles by 2035, which is not as extreme,
  but still a sizable fraction of road vehicles \cite{IHSstudy}.} In
these vehicles, the users would 
step in, select a destination, and would then be free to read, sleep,
watch a movie, answer emails, or otherwise occupy themselves without
needing to pay any attention to the operation of the vehicle. While this
vision has its benefits, it makes many people nervous about
ceding their driving agency to a computer system, especially on such a
potentially short timescale \cite{clytton}. Google's Chris Urmson is
still claiming that self-driving cars will be available in 5 years (a
claim first made in August, 2014, a personal goal based on when his
eldest son can get his license \cite{gomesObstacles}), and expects
these vehicles not to have steering wheels or the possibility of human
control. That these vehicles would operate on public roadways is not
stated explicitly---and seems optimistic given the speed of regulation
alone---but is implied in these kinds of statements. Elon Musk's
electric car company, Tesla Motors, is working on an 
``autopilot''\footnote{This name explicitly invokes aircraft
  automation, and repeats the name of Chrysler's early ``Auto Pilot''
cruise-control system, from the late 1950s
\cite{kesslerElon}. Note that automated driving will not be available
in residential neighborhoods because Tesla does not believe it is safe
enough \cite{slashgearDaviesT}.} mode for their cars, intended to be released publicly by
the summer of 2015.
Though Tesla is clear that human involvement and judgment will be
required at first, Musk suggests 
that once sufficiently capable vehicles 
come to market, human drivers may be
outlawed \cite{kingGoogle}.
In this vision, the technology is ready, or almost ready, and law and policy need to
adapt. ``Robots can already outdrive humans,'' the optimistic view says, ``now
everyone needs to get out of the way''
\cite{fisherInside}.

%%TODO (DONE)
%% ???-To be written
%% Tesla and forthcoming ``autopilot'' mode, which explicitly involved aircraft
%% discourse and refers back to the first cruise control Chrysler ``Auto Pilot''


%% 2) deal in general with some of the popular portrayals of the
%% technology:  (4 p)
%% ---you would be excused for thinking XYZ, based on what you read in
%% the popular press
%% ---but this is a flawed picture of the future, and an incomplete
%% picture of the tech

However, another common viewpoint is that major problems remain to be
solved, and that fully automated vehicle technology is not yet ready for
commercialization, whatever Urmson and Google 
say. Questions of safety, legality, and insurance cast doubts on the
optimistic timescale. Moral and ethical
quandaries---driverless car ``trolley problems''---question how
autonomous machines will respond to unusual situations and whether
people will accept those responses. Would American freedom-rhetoric
allow the outlawing of human drivers \cite{badgerElon}? 
Perhaps not. Would we accept a computerized device that kills ``3,300
Americans'' per year, but replaces the ``roughly 33,000 lives a year
that perish on U.S. roads as a result of human
error'' \cite{mcfarlandDriverless}?
This is a difficult empirical question about the propensity of people
to want to have a recognizable site for blame, and the tolerance of
human beings for computer error. Among others, Bill Gurley, an Uber
investor interviewed because of his connection to companies engaged in
driverless vehicle research, believes that given the
catastrophic potential for errors, the reliability requirement
for these systems---``four nines,'' as he describes it---renders the
technology ``a long way 
off'' \cite{shontellTop}. John Leonard, notable as one of the researchers who
worked on MIT's entrant to the DARPA Urban Challenge in 2007, has gone
on the record multiple times about his doubts, and has become a very
public face for driverless car skepticism \cite{ramseyWhen} \cite{gomesCircles}.

%% From all this, one might be excused for thinking this was simply an
%% argument between the technophiles and the naysayers, the daring dreamers versus the
%% conservative pragmatists. But 

%% Both these narratives 
%% tend to take a certain possible future of technology and either support it
%% or illustrate its problems, without challenging that fundamental
%% picture. 

%% And few people seem to dispute what the object is that is
%% being discussed, or how else it might look.

There is much more to the story than is
apparent in this simplistic dual-narrative.  A few
articles about automated vehicles remember the history, from the 1939
World's Fair \cite{CBSPetersen} to the Firebird concept cars
\cite{walshVs} to Dickmanns's work in the 
1990s \cite{HCRIDriverless} and DARPA's investments in automated
driving---the latter largely stripped of its
 military valences \cite{economistSR}. But these are generally subsumed by an imagined
 historical through-line, and both popular visions of this
technology focus on the future, and do relatively little to
investigate the past or present: how we got to 
where we are now, or even what level of autonomy we currently see in
our vehicles.\footnote{Though it should
  be said that John Leonard, for example, has much more nuanced views
  on the subject than get represented in these narratives.} The idea that ``if you are asked to take control, we've
failed'' has become deeply ingrained in our development
culture, more than just as a risk-mitigation strategy but as a guiding
ideology. But as we will see, this view does not hold together given
historical examples.

%---The future doesn't look like we think it does, pull material from
%CMS801 final paper! 
%3) and shift to the lead-in to the main thesis sections: 

%%TODO (DONE) consider: ; or even to some extent ontological position of a
%%self-driving car that ``knows'' nothing 


%% 1) What are the stories we are hearing? 
%% 2) Where do they come from?
%% 3) What are the stakes for these stories?
%% 4) What are other ones out there, why aren't they so heard? and 
%% 5) Where do we go from here? What understandings need to change to
%% make the broader picture comprehensible?

%TODO (DONE) alter the end of this if the conclusion becomes two chapters
\section{Overview}

%% The nature of the self-driving car as it will appear in
%% the near future is not self-evident, however, no matter how loud Google's voice
%% or how large their investment. These technologies and visions of them
%% enact certain ideologically charged activities---such as supporting
%% particular research agendas---and these can be traced through the
%% histories and current discourses of the fields involved.
%% The particular epistemologies of the research fields that contribute
%% to this vision greatly affect what is considered, and valued, and
%% pushed forward. And recognizing the ideological charges behind
%% self-driving vehicle 
%% work, and examining alternatives, might lead us to expect a very
%% different technology with very
%% different social and cultural impacts. 

%% I argue that 

%% This thesis is organized around several key questions. First, what are
%% the dominant narratives of self-driving vehicles and where do they
%% come from?

%% , looking
%% both at some of the literature in the field of human supervisory
%% control, and historical examples from other regimes of automation
%% that

%% Third, what other potential
%% narratives for automation,
%% particularly vehicle automation, exist?

%% Finally, where can we go, as a
%% culture, from this realization? What understandings need to change to
%% make this picture more broadly comprehensible? And what should it mean
%% for policy? While this is a broad topic, itself worthy of a thesis,

%% I
%% argue in the conclusion that the contrasts
%% between the rhetorics of artificial intelligence and supervisory
%% control, which really describe many of the same artifacts, point the
%% way.

%% Taking seriously human beings as cyborg
%% entities can alter public
%% opinion and policy in
%% productive ways.

%%  and where it is headed in the
%% future, which leads to

This thesis investigates dominant narratives of
self-driving vehicles and their gaps, and assesses the importance of
alternative paradigms or narratives of automation. In chapter \ref{chap:1}, I continue examining
the sources of the idea, common both among the media and among
self-driving vehicle researchers, that complete vehicle autonomy is
the only future vision worth discussing and investigating. And I
describe historical automation precedents that bear on automated cars. In chapter
\ref{chap:2}, I address design fictions and the issues presented by technologies involved in
automation (and the fields with which visions of full automation are
intertwined: computer vision, urban planning, machine ethics, and
others). Taking the popular narrative at face value presents the
possibility of certain kinds of futures---with important
disadvantages---while foreclosing on others in ways that must be
carefully examined. In chapter \ref{chap:3}, I
describe an alternative paradigm for thinking about automation, and
make the case for why it is an acceptable---and in
many ways, more apt---alternative for considering what automation is
and does. As I describe, supervisory control or joint-cognitive
systems research presents a very
different read on what automation means, and allows for the possibility of different futures with
different stakes and trade-offs. Finally, in the conclusion, I briefly examine the
historical re-shaping of roadways, and argue that adding autonomy to cars does not,
alone, solve any social problems. Successfully addressing social
issues requires careful design at a whole-system level, involving
people, vehicles, infrastructure, and a broader appreciation for our hybrid
engagements with machines. This appreciation allows us to shift the fundamental
questions we ask about technology in a way that is, hopefully,
productive for research and policy-making.

%% The deeper one 
%% goes, the more ambivalent the mess becomes. Certain ideas seem
%% constant across projects, but others differ wildly. Their stated goals
%% often seem subtly off-set from their assumptions and end-products. And
%% even more fundamentally, the idea of the ``self-driving'' vehicle,
%% where the human does nothing at all, is hardly a self-evident one.

%% \footnote{The time has not yet come when the social
%% and cultural impacts can reasonably be predicted in a cohesive way.
%% But, as I think such things are absolutely fundamental to the way we
%% should think about new technologies, I am unwilling to drop them
%% entirely. So what follows is a patchwork, though, I hope, an artful
%% one.} Different technological possibilities produce several visions of
%% what society might be, but e

%% something
%% understood and on which a meaningful ideological critique could be
%% based

I originally set out to critically examine the cultural and social
implications of self-driving vehicle technology. In my na\"{\i}vet\'{e}, I
conceived of driverless cars as just-another-AI-technology. But as I
researched, I discovered that not only did 
self-driving cars lack a fixed social meaning---their uses and
contexts still up for grabs---but the technology itself is unspecified
in profound and important ways. In an apt lesson for an early researcher,
here was a thing that had not reached the stage of becoming, in
Latour's parlance, a ``black box,'' despite the media's treatment of
it as such \cite{LatourScience}. Below the surface, controversy
bubbles. Turning up other research paths---and indeed, numerous historical
counterexamples---made it painfully clear that the story I wanted was
not the story I could tell. Each social vision is
contingent, and will ultimately have to come into
contact not only with physical reality, but the social realms of law,
technology, and public acceptance. My research materials have
therefore consisted not only of public media portrayals of automated
vehicle efforts---comprising hundreds of news articles accessed
between March 2014 and April 2015---but also other public
documentation including patents; research papers in robotics,
computer science, law, and ethics; and books on computers, automation systems,
and their human 
interrelationships. I attended three information-technology or urban-planning
related conferences at MIT to examine the presentation of automated
vehicles within technical communities. I have also conducted a handful
of one-hour interviews with 
automated vehicle and transport researchers to attempt to gain
internal perspective on their concerns and motivations. Rather than
projecting from visions to cultural effects, I have had to dig deeper:
putting these visions in the context of 
historical fields and precedents, exposing their gaps. And I hope this project has become a  
deep look into the ideologies and rhetorics of the self-driving car,
one which hints at what society might be like once these
vehicles are common, but more so investigates what different models for these vehicles
exist, and what their public face might be if different people wrote,
and different ideas guided, their stories.

Fundamentally, I am a scientist at heart,\footnote{I mean this not necessarily in
  the methods of what is considered modern, experimental science,
  which does not fit easily within the realms of historical critique,
  but in terms of ethos, and some notion of falsifiability.} and I
seek to approach my 
work in that light. What idea, or hypothesis, does this thesis attempt
to disprove? While I say many things that have significant backing,
though they may not themselves be provable\footnote{I hope they are
  convincing, or at least suggestive.} in any real sense, my core case
is that there is more than one story to tell about road vehicle
automation. One can argue about approaches to automation and autonomy,
based on historical examples, empirical data, or ethnographic
findings, but the idea that the conventional story about automation
(that it is clearly and naturally headed toward completeness and
totality) is right, true, and the only story to tell is clearly
without basis. Ultimately, I argue that the narrative of self-driving cars, as
generally conceived, is far narrower than history suggests it should
be. A broader view of this narrative shows a different side of
automated vehicles that is more hybrid, more complex, and presents
very different questions for policymakers, engineers, designers, and
the public. Stories have power, and it
behooves me, on finding the stories we tell to be so deeply one-sided,
to try to rebalance the narrative.

%% Note: I need more voices from actual modern driverless car research
%% papers, though looking at the Dickmanns ones more is a good idea! need
%% a section specifically on how the ideologies come through!

